Using an SMT engine to generate Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-13 2017-Xudong-SMT/ActaInformatica.tex
inserted Enable example

In Fig.~\ref{schema:ArchFailure:pNet}, we show examples of these
constructs, with two pLTSs, one hole and one pNet node encoding our running example.

\section{Running example}
\label{section:FailureTimerMax}
\label{section:BIParchitectures}
\label{section:examples}

%% Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
%% many examples of pNets, encoding
%% operators of various classical process algebras, or more complex
%% synchronization structures in distributed or parallel languages.

%% \TODO{1.5 page including figures ?}

%% \TODO{Simon: can you do this in less than one page ? If you prefer,
%%   merge with ``Running example'' may be easier to describe the syntax}

%% \TODO{Simon: or may be better, present BIP architectures and the use-case \_BEFORE\_ The pNets definition, and define the encoding here}

%% \TODO{Should have here both the original BIP drawing (Simon), and the
%%   pNet encoding (fig by Eric)}

@@

\begin{example}[An open transition]
  This transition is generated by application of the vector \texttt{SV0},
  synchronizing the initial actions of pLTSs \texttt{Control} and
  \texttt{Timer} (see Fig.~\ref{schema:ArchFailure:pNet}), with an action of the hole \texttt{B} equal to
  \texttt{fail}. The local variable \texttt{t} of the \texttt{Timer}
  is assigned to its initial value \texttt{Max}.
 A full output of 
the use-case is provided in~\cite{Avocs-RR}.

  \label{OT:example-1}
        \begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s0~{\xrightarrow{\mathit{fail}(\true)}} ~s1, t0~{\xrightarrow{\mathit{start}(\true)}} ~t1\}
	  \{\xrightarrow{\mathit{hb}}\}, \mathit{hb}=\mathit{fail}(\true) \land v=\underline{\mathit{fail}}, \{t := \mathit{Max}\}
        }
	{\ostate{s0,t0} \xrightarrow {v} \ostate{s1,t1}}
	\end{mathpar}
\end{example}
%
%The open transition $ot_2$ involves the possible actions from the process $P$ and prefix expression $acc(x);Q$. The hole behaviour, represented by variable $hb_{12}$ must be equal to the exit action $\delta(x)$, and the accept event $acc(x)$ will capture the same value $x$.
%
%The transition in the conclusion of this rule starts in
%state $\ostate{00}$, where both $C_1$ and $C_2$ are in their initial
%states. It ends in a different structured state, because in $ot_2$
%both pLTS conduct transitions, showing that the control has gone from
%processes $P$ to $(acc(x);Q$, consuming actions $\delta(x)$ and
%$acc(x)$. Controller $C_1$ has only one state, but its state variable
%$s_0$ has changed to 1.

@@

\TODO{the proof of the finiteness theorem refers to SOS rules Tr1 and
  Tr2. So maybe we must include (and explain) these here. One problem
  is that these rules have been modified/improved with each new
  publication of pNets papers, so we have to decide which version we
  use here}

@@

\begin{proof}
  \TODO{I tentatively copied the proof here, from RR-8898. We should
    check if it is sufficient with this version of the algorithm, or
    if it can be make more precise (some added analysis of the
    behavior of matching/combining/filetring ?)}
  The possible set of states of the open-automaton is the cartesian
  product of the states of its leaves $pLTS_i^{i\in L}$, that is finite by
  hypothesis. So the top-level residual loop of Algorithm 1 terminates provided each 
  iteration terminates. The enumeration of open-transitions in step 2b is bounded by the
  number of applications of rules Tr2 on the structure of the
  pNet tree, with a finite number of synchronisation vectors applying
  at each node the number of global open transition is finite. Similarily rule Tr1 is 
  applied finitely if the number
  of transitions of each pLTS is finite. So we get finitely many deduction trees, and
  open-transitions which ensures that each internal loop of Algorithm 1 terminates.
  \end{proof}

@@

\begin{figure}[t]
\begin{mathpar}
    ot = \openrule{\{s0 \xrightarrow{fail(true)} s0, t0 \xrightarrow{resume(false)} t0\}, ~~\{\xrightarrow{hb}\},~~fail(true)=fail(b_1)\\
    \OTland resume(false)=start(b_2)\OTland hb=fail(b_0)\OTland v=\underline{fail}\OTland b_1=b_2\\
    \OTland (b_1\lor b_2) \Rightarrow b_0, ~~\{\}}
    	   {\ostate{s0,t0} \xrightarrow{v} \ostate{s0,t0}}
  \end{mathpar}
\caption{One of the unsatisfiable open transitions in the Failure Monitor pNet}  \label{schema:unsat-ot}
\end{figure}

@@

\TODO{Candidate theorems:
  - All reachable states from algorithm 1 are kept by Algorithm 3\\
  - Algorithm 3 computes an open automaton that is minimal with
  respect of SAT checking decidability.\\
  Of course we have to define this notion of minimality, that may be a
  little tricky}

@@

\begin{figure}[t]
  %  \centerline{\includegraphics[width=12cm]{XFIG/SMTLIB2}}
    \centerline{\includegraphics[width=\linewidth]{XFIG/SMTLIB3}}
  \caption{The input of the Z3 solver in SMT-LIB language and the output result}  \label{schema:smt-lib}
\end{figure}

@@

%% \noteSB{I am not sure I understand the entire message of this phrase}\hl{though in practice our tool only needs the brute information ``sat''
%% or ``unsat''.
%% The result "unsat" in the output is just what we expected.
%% If the result is ``sat'', and more details are needed, Z3 can exhibit a
%% model (i.e. an instanciation of the variables) showing the satisfiability.
% as appears in the ``unsat'' proof generated by Z3 in which ``\{= ?x66 ?x70\}'' is shown to be false (see Fig. \ref{schema:smt-lib})

@@

\section{Use-cases}
\TODO{new section here, showing various features on different examples. Possible: Enable operator in Lotos, with (equivalent) state-based and data-based encodings; Associativity of Enable; simple communication protocol, specification and implementation; BIP FailureTimer architecture, including non-trivial ``Interval'' data-type and operators; without and with encoding of maximal progress.}

\subsection{Result for the running example}
\label{section:resultOA}
\TODO{This is the original example from the Avocs paper}
For this example, the tool builds 184 open transitions, whereof 173
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 11 open
transitions, is shown in Fig.~\ref{schema:resultOA}.

@@

\subsection{Extended version of the FailureTimer BIP Architecture}
\label{section:resultTimerInterval}
\TODO{Here we have extended the FailureTimer architecture using time intervals,
  that makes more interesting the timeout/reset temporal behavior}

We present here an extended version of the FailureTimer BIP
architecture, in which the timeout action can occur
non-deterministicaly during some time interval. This behavior is more
realistic, but ensuring safety is a bit more complicated, as it
involves reasonning on the time intervals in the predicates of the
open automaton.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{XFIG/BIPspec-ArchFailureTimerMax-v4}
  \caption{The BIP specification of the Failure Monitor architecture}
  \label{fig:schema:ArchFailure:BIP}
\end{figure}

Figure \ref{fig:schema:ArchFailure:BIP} shows components
$\NameTimer$(imer) and $\NameCtrl$(ontrol), with interfaces
$\btuple{
  \{t, z\},
  \{\PortTick, \PortCancel, \PortTOT\},
  \bigl\{\!\PortTick \mapsto \{t, z\}\!\bigr\}
}$,
and
$\btuple{
  \{\mathit{zone}\},
  \{\PortReset, \PortFail, \PortResume, \PortTOC\},
  \bigl\{\!\PortFail \mapsto \{\mathit{zone}\}\!\bigr\}
}$
respectively.
Variable $t$ is implicitly assumed to be an integer.  Variables $z$
and $\mathit{zone}$ are  integer intervals of
the form $[l,u]$.

\TODO{refer to Research Report (not yet avaialble) or add an appendix,
  for a full explanation of this example, 
  and the translation from BIP to pNets}

Figure \ref{fig:FailureTimer:pNet} shows the encoding of the Failure Monitor
architecture, including the encodings of the two coordinators, \ie
$\nopri[\NameTimer]$ and $\nopri[\NameCtrl]$.

>>>>

In Fig.~\ref{schema:EnableEncodings}, we show examples of these
constructs, with two pLTSs, one hole and one pNet node encoding our running example.

\subsection{Example}
\label{section:example}
\TODO{Taken from \cite[Forte]}

\begin{figure}[t]
\begin{minipage}{6.1cm}

  \includegraphics[width=\linewidth]{XFIG/Enable1}
  \\[1.3ex]
 \includegraphics[width=\linewidth]{XFIG/Enable2}
  \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
\end{minipage}
  \hspace{2mm}
\begin{minipage}{6cm}
  \includegraphics[width=\linewidth]{XFIG/P-QR}
  \caption{Composed pNet for ``P>>(Q>>R)''}  \label{schema:enable-composed}
\end{minipage}

\end{figure}

	To give simple intuitions of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In Fig. \ref{schema:enable-pnets}, we show 2 possible encodings of the  
	``Enable'' operator.
	In the Enable expression ``P>>Q'', an \texttt{exit(x)} statement within P
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept(x)} statement of Q.

	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit(x)} and \texttt{accept(x)}. Both $\delta(x)$
	and $acc(x)$ actions are
	implicitly included in the sorts of all processes.
	We need no specific predicate over the action expressions, apart
	from equality of actions. 
	In the first encoding $Enable1$, in the upper part of
	Fig. \ref{schema:enable-pnets}, we use a controller $C_1$ with two
	states, and simple control actions $l, r, \delta$. The second encoding
	$Enable2$ uses a data-oriented style, with a single state controller, and a
	state-variable $s_0$, with values in $\{0,1\}$.

        In this example we use \emph{synchronised
          actions}, to express actions that cannot be further synchronised, like the
        $\tau$ silent action of CCS. We denote them as any action expression with the text
        underlined, as e.g. $\underline{\delta(x_2)}$. Such
        synchronised actions do not play any special role for defining
        the semantics of pNets, but as one can expect, will be crucial
        for defining weak equivalences.

	Note that synchronisation vectors are defined in a parameterised
	manner:  the first and third lines represent one vector
	for each parameterised action in the Sort of hole $P$
	(resp. $Q$). This notation can also use predicates, as in the first
	case, in which we want the vector to apply to any action of $P$
	except $\delta(z)$.

	In Fig. \ref{schema:enable-composed}, we enrich our example by
	composing 2 occurences of the Enable1 pNet. To simplify,
	we only have represented one instance of the synchronisation vector set,
	and of the controller.

@@

\begin{example}[An open transition]\TODO{change for Enable example}
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $C_3$ has
    changed from state 0 to 1.

    \TODO{add another example to illustrate assignements/Post}

  \label{OT:example-1}
        \begin{mathpar}
	  OT_1  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [\forall z. acc(x3) \neq \delta(z)]
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{1,0}}
	\end{mathpar}
\end{example}

@@

@@

\begin{proof}
  \TODO{I tentatively copied the proof here, from RR-8898.
the proof of the finiteness theorem refers to SOS rules Tr1 and
  Tr2. So maybe we must include (and explain) these here. One problem
  is that these rules have been modified/improved with each new
  publication of pNets papers, so we have to decide which version we
  use here
    Then we should
    check if it is sufficient with this version of the algorithm, or
    if it can be make more precise (some added analysis of the
    behavior of matching/combining/filetring ?)}
  The possible set of states of the open-automaton is the cartesian
  product of the states of its leaves $pLTS_i^{i\in L}$, that is finite by
  hypothesis. So the top-level residual loop of Algorithm 1 terminates provided each 
  iteration terminates. The enumeration of open-transitions in step 2b is bounded by the
  number of applications of rules Tr2 on the structure of the
  pNet tree, with a finite number of synchronisation vectors applying
  at each node the number of global open transition is finite. Similarily rule Tr1 is 
  applied finitely if the number
  of transitions of each pLTS is finite. So we get finitely many deduction trees, and
  open-transitions which ensures that each internal loop of Algorithm 1 terminates.
  \end{proof}

@@

\begin{figure}[t]
\begin{mathpar}
    ot = \openrule{\{s0 \xrightarrow{fail(true)} s0, t0 \xrightarrow{resume(false)} t0\}, ~~\{\xrightarrow{hb}\},~~fail(true)=fail(b_1)\\
    \OTland resume(false)=start(b_2)\OTland hb=fail(b_0)\OTland v=\underline{fail}\OTland b_1=b_2\\
    \OTland (b_1\lor b_2) \Rightarrow b_0, ~~\{\}}
    	   {\ostate{s0,t0} \xrightarrow{v} \ostate{s0,t0}}
  \end{mathpar}
\caption{One of the unsatisfiable open transitions \TODO{to be changed}}  \label{schema:unsat-ot}
\end{figure}

@@

\TODO{Candidate theorems:
  - All reachable states from algorithm 1 are kept by Algorithm 3\\
  - Algorithm 3 computes an open automaton that is minimal with
  respect of SAT checking decidability.\\
  I have problem with minimality... Because of the ``two phases'' SAT
  checking that we have now, because we cannot use incomplete sets of
  variable assignments during the ``smart'' phase, and use them later,
minimality becoomes trivialy equivalent in the original and smart algorithms...}

@@

\begin{figure}[t]
  %  \centerline{\includegraphics[width=12cm]{XFIG/SMTLIB2}}
    \centerline{\includegraphics[width=\linewidth]{XFIG/SMTLIB3}}
  \caption{The input of the Z3 solver in SMT-LIB language and the output result\TODO{Change}}  \label{schema:smt-lib}
\end{figure}

@@

@@

\section{Use-cases}
\TODO{new section here, showing various features on different examples}

\subsection{Enable operator} in Lotos, with (equivalent) state-based and data-based encodings; Associativity of Enable;

\subsection{simple communication protocol} specification and implementation;

\subsection{BIP FailureTimer architecture}

\label{section:FailureTimerMax}
\label{section:BIParchitectures}
\label{section:examples}

\paragraph{Variations on the FailureTimer example}

\TODO{This is the original example from the Avocs paper, connectors slightly adapted so that it can be composed. Then Max Progress may be required to ensure properties}

For this example, the tool builds 184 open transitions, whereof 173
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 11 open
transitions, is shown in Fig.~\ref{schema:resultOA}.

@@

\subsection{Composing 2 FailureTimer BIP Architecture}
\label{section:resultTimerInterval}
\TODO{Enhanced FailureTimer, 2 instances composed, with duplicated control and Timer}
